---
title: "readall"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{readall}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
#library(readall)
```

The package `readall` offers a single interface in order to read
various types of data files. The following data file types are supported:

- **FWF** (which stands for **Fixed Width Files**): The data is stored
  in a text file (**ANSI** or **UTF-8**) and each data column has a fixed
  width (fixed number of bytes).
- **DSV** (wich stands for **Delimiter Separated Values**): The data is stored
  in a text file (**ANSI** or **UTF-8**) and the data columns are
  separated by a delimiter symbole (usually `","` or `";"`).
- **EXCEL**: Excel files (`*.xlsx` or `*.xlsm`)
- **SAS**: SAS data files (`*.sas7bdat` or `*.sas7bcat`)

## Installation

```{r eval = FALSE}
devtools::install_github('a-maldet/readall', build_vignettes = TRUE)
```

## Usage

### About `file_structure` class objects

In order to read data files, which are not stored as an R data file, we often
have to supply additional information about the file structure to the 
reading operation (column names, column types, delimiter symbols etc.).
Often multiple files share the same structure. Therefore, the `readall`
package offers the `file_structure` class, which can be
used in order to store all information about the structure of a
data file in a single `file_structure` class object.
This can be done by using one of the following functions:

- `new_file_structure_fwf()`: Define the file structure of an FWF data file
- `new_file_structure_dsv()`: Define the file structure of a DSV data file
- `new_file_structure_excel()`: Define the file structure of an EXCEL data file 
- `new_file_structure_sas()`: Define the file structure of a SAS data file

The created `file_structure` class object can be used in order to read
all data files which share the defined structure.


### About `file_definition` class objects

In the section before, we defined `file_structure` class objects, which 
hold the file structure information. This type of information
can be valid for several data files. But there are also informations,
which are only valid for a single data file. For example the file path to
the data file.
Therefore, the `readall` package offers a `file_definition` class, which
extends the `file_structure` class. This means that
a `file_definition` class object contains all needed information about the
file structure information and additionally some information that is
only valid for a specific file (like the file path).
A `file_definition` class object holds all information, which is neccessary
in order to read a specific data file with the function `read_data()`.
The following functions can be used in order to create a `file_definition`
class object:

- `new_file_definition()`: With this function you can use an existing
  `file_structure` class object in order to create a `file_definition` class
  object, by appending all needed file specific information to the
  given `file_structure` object.
  Depending on the file type defined in the given `file_structure` class object
  the resulting `file_definition` class object can describe FWF, DSV, EXCEL
  or SAS data files.
- `new_file_definition_fwf()`: Create a `file_definition` class object for FWF data files
- `new_file_definition_dsv()`: Create a `file_definition` class object for DSV data files
- `new_file_definition_excel()`: Create a `file_definition` class object for EXCEL data files
- `new_file_definition_sas()`: Create a `file_definition` class object for SAS data files

### About `file_collection` class objects

Sometimes, it is neccessary not only to read a single data file, but to read a
collection of data files and concatenate all data sets into a
single data frame. Sometimes, the data files in such a file collection
can contain data files of different file structure or even file types (for example
a mix of SAS, EXCEL, FWF and DSV data files).
For this case, the `readall` package offers a `file_collection` class. 
In a `file_collection` class object, you can store multiple `file_definition`
class objects, which define the needed data files. When calling the function
`read_data()`, all defined data files will be read an automatically 
concatenated into a single data.frame.
The function `new_file_collection()` is used in order to create such 
`file_collection` class objects.

### About adapter functions

When reading a collection of data files,
it is often neccessary to post process the each data.frame,
before concatenating all data.frames.
E.g. recode some variable levels, calculate new columns or rename
existing columns.
For this reason, the `readall` package offers adapter functions, which
are functions of the typ `f: DATA.FRAME -> DATA.FRAME`. This means,
that an adapter function takes a `data.frame` and returns a mutated version
of the `data.frame`.
An example of an adapter function would be the following function:

```{r eval = FALSE}
g <- function(x) {
  x[x$a > 1,]
}
```

Usually, on does not only want to perform a single data transformation, but
an entire list of data transformations. For this reason the `readall` package
offers the function `new_adapters()`, which can be used to store multiple
adapter funcitons in a single `adapters` class object. An `adapters` 
class object is a list of adapter functions, which can be stored in a
`file_structure` or a `file_definition` class object.
For example:

```{r eval = FALSE}
structure_excel_1 <- new_file_structure_excel(
  col_names = c("city", "adult"),
  col_types = c("character", "logical"),
  adapters = new_adapters(
    function(x) {
      names(x) <- c("CITY", "ADULT")
      x
    },
    function(x) {
      x <- x[!is.na(x$CITY),]
      x$CITY <- trimws(x$CITY)
      x
    }
  )
)
```

When calling `read_data()` the data will be read from the EXCEL file and
then the columns will automatically be renamed to `CITY` and `ADULT` and 
all rows with missing values for `CITY` will be removed. Furthermore,
the strings in `CITY` will stripped of all leading and trailing white spaces.

### About meta data

The package `readall` allows you to add meta data to `file_structure`
and `file_definition` class objects. This meta data can contain a detailed
description of each column and its value leves.
The meta data of a single data column is stored in a `col_meta` class object,
which can be created with the command `new_col_meta()`.
The meta information for all columns can be collected with the command
`new_file_meta()`.

Example-1:
```{r eval = FALSE}
structure_excel_1 <- new_file_structure_excel(
  col_names = c("city", "adult"),
  col_types = c("character", "logical"),
  meta_list = new_file_meta(
    new_col_meta(
      desc = "National and international city codes",
      values = c("XXXX", "A", NA),
      values_desc = c("4 digit city code", "abroad", "missing")
    ),
    new_col_meta(
      desc = "Is the person an adult",
      values = c(TRUE, FALSE, NA),
      values_desc = c("The person is an adult", "The person is a child", "unknown")
    )
  )
)
```

Example-2:
```{r eval = FALSE}
structure_excel_1 <- new_file_structure_excel(
  cols = list(
    list(
      name = "city",
      type = "character",
      new_col_meta(
        desc = "National and international city codes",
        values = c("XXXX", "A", NA),
        values_desc = c("4 digit city code", "abroad", "missing")
      )
    ),
    list(
      name = "adult",
      type = "logical",
      new_col_meta(
        desc = "Is the person an adult",
        values = c(TRUE, FALSE, NA),
        values_desc = c("The person is an adult", "The person is a child", "unknown")
      )
    )
  )
)
```

If a data file is read with the command `read_data()`, then all
specified meta informations will be added to specific columns of the resulting
`data.frame`.

The meta data stored in a `file_structure` or a `file_definition` class object
or a `data.frame` generated by calling `read_data()` can be extracted by
calling `get_meta()`.

Example-3:
```{r eval = FALSE}
df_meta <- get_meta(file_definition_1, cols = c("city", "adult"))
```

Example-4:
```{r eval = FALSE}
data <- read_data(file_definition_1)
df_meta <- get_meta(data, cols = c("city", "adult"))
```

### Important commands

The following section describes the most important function of `readall`.
Each function is documented and the documentation of each function can be
displayed by calling `?readall::FUNCTIONNAME`.

#### Create `file_structure` class objects

`file_structure` class objects contain all information about the file
structure, which can be valid for several data files (e.g. column structure,
delimiter symbols, column names, data types etc.). This class objects
can be created with the following commands:

- `new_file_structure_fwf()`: File structure of FWF data files
- `new_file_structure_dsv()`: File structure of DSV files
- `new_file_structure_excel()`: File structure of EXCEL files
- `new_file_structure_sas()`: File structure of SAS files

Example:

```{r eval = FALSE}
structure_fwf_1 <- new_file_structure_fwf(
  cols = list(
    list(
      type = "character",
      name = "sex",
      start = 1
    ),
    list(
      type = "numeric",
      name = "age",
      start = 3
    ),
    list(
      typ = "numeric",
      name = "city",
      start = 7
      
    )
  ),
  sep_width = 1,
  adapters = new_adapters(
    function(x) {
      x$sex <- ifelse(x$sex == "m", "male", "female")
      x
    }
  )
)
```

The code above creates a `file_structure` class object for FWF data files.
This data files contain three column (sex, age and city codes).
The `sex` column starts with the first row character and contains a single
character. After that, there is a blank space (`sep_width = 1`) and then
cames the `age` column consisting out of three characters. After that
is again a blank space and finally comes the `city` column holding the
city codes. The created `file_structure` class object contains an
adapter function, which will automatically be executed after reading a
data file with this `file_structure` object. The adapter function
recodes the `sex` column to full length strings.

#### Create `file_definition` class objects

A `file_definition` class object contains all information can also be
stored in `file_structure` class objects (information about the file structure
which can be valid for all files of the same structure),
but it also contains information that is only valid for a single specific
data file. `file_definition` class objects can be created with the following
command:

- `new_file_definition()`: Takes an existing `file_structure` class objects and
  extends it to a `file_definition` class object.
  Depending on the `file_structure` class object the resulting `file_definition`
  class object can be describe FWF, DSV, EXCEL and SAS data files.
- `new_file_definition_fwf()`: Creates a `file_definition` class object for FWF data files
- `new_file_definition_dsv()`: Creates a `file_definition` class object for DSV data files
- `new_file_definition_excel()`: Creates a `file_definition` class object for EXCEL data files
- `new_file_definition_sas()`: Creates a `file_definition` class object for SAS data files

Example:

```{r eval = FALSE}
file_definition_file_1 <- new_file_definition(
  file_path = "C:/file1.dat",
  file_structure = structure_fwf_1,
  extra_adapters = new_adapters(
    function(x) {
      x[x$age >= 30 && x$age < 40,]
    }
  )
)
```

The created `file_definition` class object `file_definition_file_1`
extends the `file_structure` class object `structure_fwf_1`, which was defined
earlier.
Therefore, it is a `file_definition` for an FWF data file, which is 
located at `C:/file1.dat`. After reading this data file, the read data set is
automatically filtered by age, such that only persons with age between 30 and
40 are kept.

#### Create `file_collection` class objects

A `file_collection` class object contains a list of `file_definition`
class objects, describing different data files, whose data should
automatically be concatenated after reading the data files.
A `file_collection` class object can be created with the command
`new_file_collection()`.

Example:

```{r eval = FALSE}
file_collection_1 <- new_file_collection(
  file_definition_file_1,
  file_definition_file_2,
  file_definition_file_3,
  cols_keep = c("sex", "age"),
  extra_adapters = new_adapters(
    function(x) {
      x[x$sex == "male",]
    }
  )
  extra_col_file_path = "file"
)
```

The `file_collection` class object in this example contains the 
informations of 3 different data files. If `read_data()` is applied on
`file_collection_1`, then the following steps are executed automatically:

1. All 3 data files are read and each data set is stored in a `data.frame`
2. For each `data.frame` only the columns `sex` and `age` are kept
3. The column `file` is added to each `data.frame`, holding the file path of each data file
4. The adapter functions are executed on each `data.frame`.
   For the data file defined in `file_definition_1`, three adapter functions
   were defined. First, the column `sex` is recoded, then the `data.frame` is
   filtered by `age` and finally only entries with `sex == "male"` are kept.
   For the data files defined in `file_definition_2` and `file_definition_3`
   we don't know if there were adapter functions defined before 
   (when the `file_definition` class objects were defined), but 
   for both data files there is at least one adapter function defined, which
   filters `sex == "male"`.
5. All three `data.frames` are concatenated into a single `data.frame`

The resulting `data.frame` could look something like this

```{r eval = TRUE, echo = FALSE}
data.frame(
  sex = rep("male", 7),
  age = c(16, 67, 31, 84, 47, 33, 98),
  file = c("R:/daten/file1.dat", "R:/daten/file1.dat", "R:/daten/file2.sas", "R:/daten/file2.sas", "R:/daten/file3.excel", "R:/daten/file3.excel", "R:/daten/file3.excel")
)
```

#### Reading single data files

The following functions can be used, in order to read data files:

- `read_data()`: Read the FWF, DSV, EXCEL or SAS data file, which was defined
  in the passed in `file_definition` class object.
- `read_data_fwf()`: Read an FWF data file. Instead of passing in a
  `file_definition` class object holding all needed informations,
  the needed file definitions are passed directy to the function. 
- `read_data_dsv()`: Read a DSV data file. Instead of passing in a
  `file_definition` class object holding all needed informations,
  the needed file definitions are passed directy to the function. 
- `read_data_excel()`: Read an EXCEL data file. Instead of passing in a
  `file_definition` class object holding all needed informations,
  the needed file definitions are passed directy to the function. 
- `read_data_sas()`: Read a SAS data file. Instead of passing in a
  `file_definition` class object holding all needed informations,
  the needed file definitions are passed directy to the function. 

Example-1:

```{r eval = FALSE}
df1 <- read_data(file_definition_file_1)
```

Example-2:

```{r eval = FALSE}
df2 <- read_data_fwf(
  file_path = "R:/daten/file1.dat",
  cols = list(
    list(
      type = "character",
      name = "sex",
      start = 1
    ),
    list(
      type = "numeric",
      name = "age",
      start = 3
    ),
    list(
      typ = "numeric",
      name = "city",
      start = 7
      
    )
  ),
  sep_width = 1,
  adapters = new_adapters(
    function(x) {
      x$sex <- ifelse(x$sex == "m", "male", "female")
      x
    },
    function(x) {
      x[x$city >= 70000 && x$city < 80000,]
    }
  ),
  cols_keep = c("sex", "age"),
  extra_col_file_path = "file"
)
```

#### Read multiple data files at once

The function `read_data()` does not only take `file_definition` class objects,
but also `file_collection` class objects. By using `file_collection` class
objects, we can also read multiple data files at once.

Example:

```{r eval = FALSE}
df_gesamt <- read_data(file_collection_1)
```

The command above reads all 3 data files defined in `file_collection_1`.
The following steps are executed automatically:

1. All 3 data files are read and each data set is stored in a `data.frame`
2. For each `data.frame` only the columns `sex` and `age` are kept
3. The column `file` is added to each `data.frame`, holding the file path of each data file
4. The adapter functions are executed on each `data.frame`.
   For the data file defined in `file_definition_1`, three adapter functions
   were defined. First, the column `sex` is recoded, then the `data.frame` is
   filtered by `age` and finally only entries with `sex == "male"` are kept.
   For the data files defined in `file_definition_2` and `file_definition_3`
   we don't know if there were adapter functions defined before 
   (when the `file_definition` class objects were defined), but 
   for both data files there is at least one adapter function defined, which
   filters `sex == "male"`.
5. All three `data.frames` are concatenated into a single `data.frame`

#### Read meta data

The function `get_meta()` can extract meta data from:

- `file_structure` class objects
- `file_definition` class objects
- `file_collection` class objects
- `data.frames` created by calling `read_data()` or `read_data_*()` 
  (where `*` can stands for fwf/dsv/excel/sas)

The command `get_meta()` returns a `data.frame` holding the following columns:

- `col_name`: A charcter column, holding the names for each data column
- `col_id`: A numeric column, holding the positions of each data column
- `col_type`: A character column, holding data type of each data column
- `col_desc`: A character column, describing each column in detail
- `col_values`: A character column, holding a text version the value levels of each column
- `col_values_desc`: A character column, describing each value level of each column
- `col_valid_start`: A character column, giving some information since when
  the variable is valid
- `col_valid_end`: A character column, giving some information till when
  the variable is valid
- `file_path`: A character column, holding the file paths of the data files.

#### Change existing `file_definition` or `file_collection` objects

Sometimes it is useful to modify some attributes of a `file_definition` or a 
`file_collection`. The following commands can be useful:

- `set_cols_keep()`: Set the columns which should be kept when calling `read_data()`
- `set_extra_col_file_path()`: Set the name of the column, in which
  the file path of the data files should be stored.
- `set_n_max()`: Set the maximum number of rows to be read when calling
  `read_data()`. If applied to a `file_collection` this value is set for 
  each `file_definition` contained in the collection.
- `set_adapters()`: Set the adapters attribute. This argument overwrites all
  existing adapter functions. If applied to a `file_collection`, this
  command is applied to each `file_definition` contained in the collection.
- `add_adapters()`: Append a set of adapter functions to the already defined
  adapter functions. If applied to a `file_collection`, the adapter functions
  are appended to each `file_definition` contained in the collection.

